---
title: "Data Checks Prior to Hierarchical Clustering in PCOS"
author: "Dunaif Lab"
date: "updated 6/9/2021"
output: 
  html_document:
    toc: true
    toc_depth: 3
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

This document was created to review data for potential issues prior to clustering.  The data will be looked at in a series of steps before we determine our clean set of data to use as part of clustering.  These steps are:

1. Use entire dataset.
  + This will allow us to review observations outside of technical limits of the assay as well establish a mean and standard deviation indicative of the entire cohort to help us explore potential outliers in the data set of cases that can be used for clustering.
2. Review each trait against the technical and logical ranges for each assay method.
  + These ranges will need to be input for each assay method observed in the dataset. These limits will then be used to shows which datapoints are outside of these ranges for each trait.
  + Any points outside of the technical limits should be set to the upper/lower limit of the assay's technical limits.
3. Removal of incomplete data
  + Any observations missing any of the traits required for clustering is incomplete and will be removed before the next step.
  + Before moving forward all data outside of technical and logical limits should be reviewed.
4. Review trait raw trait data.
  + Each trait will have the following tables/plots presented
    + Table with descriptive statistics by assay method
    + Histogram and scatter plot to review the distribution of data by assay method
    + Box plots: Raw trait and log(trait) values are presented.  The raw values contain points with labels for any points outside of 4SD.  The SD were calculated from the larger cohort before removal of incomplete data.  These points should be reviewed to determine if they are outliers.
    + The labeled points will also be displayed in a table.


## Data Description

This is a list of all European PCOS that we have record of trait data- either by genotyping or in current databooks. Phenotypic data was included regardless if genotyped or not. Internal subjects checked for current PCOS status. 


```{r read_data, eval=TRUE, echo=FALSE, error=FALSE, message=FALSE}
library(ggrepel)
library(tidymodels)
library(reshape2)
library(factoextra)
library(FactoMineR)
library(ClusterR)
library(fpc)
library(stats)
library(gplots)
library(clusterSim)
library(factoextra)
library(FactoMineR)
library(dplyr)
library(describedata)
library(lattice)
library(corrplot)
library(DT)

 ################################################
# Functions                                    #
################################################

## Reverse normal transformation
rntransform <- function(y) {
  out <- rank(y)
  out[is.na(y)] <- NA
  qnorm((out - 0.5) / max(out, na.rm=T))
}


################################################
# Variables                                    #
################################################

## Path/file of the input file, which must have columns according to
## cluster_input_cols below.
input_file <- "~/Documents/PCOS/Clustering/PCOS_Alltraits.txt"


## Names of sample ID, age, bmi, and the other 7 traits
sample <- 'sample_id'; age <- 'age'; bmi <- 'bmi'
traitnames = c('T', 'dheas', 'i0', 'g0', 'shbg', 'lh', 'fsh')

## variable names for displaying/plotting results
var_labels <- c('BMI', 'T', 'DHEAS', 'Ins0', 'Glu0', 'SHBG', 'LH', 'FSH')

## The assay method for each of the 7 traits is another variable, named by
## adding a postfix to the trait name.  The postfix in our data is
## '_assay_method'.  For example, fsh_assay_method.
assay <- '_assay_method'

## columns of the input file
cluster_input_cols <- c(sample, age, bmi,
                        traitnames,
                        paste0(traitnames, assay))
## columns of variables (some to be defined later) for clustering
d <- '.'
cluster_cols <- c(sample, age,
                  paste(c(bmi, traitnames), 'z', sep=d))

## Methods for distance calculation and for hierarchical clustering.
## Our initial analysis used manhattan distance and ward.D clustering method.
## ward.D2 is probably a better choice.
dist_metric <- 'manhattan'
clust_method <- 'ward.D'

# Vectors for assay method values and technical and logical ranges
# Into each vector, enter the assay identifier you used and the technical and logical ranges 
# in the same order.

# Testosterone Limits
T_assay_values <- c(1, 3, 4, 5, 6, 7)
T_assay_tech_max <- c(1600, 735, 1800, 1600, 1600, 1600)
T_assay_tech_min <- c(4, 7, 2, 4, 4, 4)
T_assay_log_max <- c(150, 150, 120, 150, 150, 150)
T_assay_log_min <- c(15, 7, 5, 20, 15, 10)
# SHBG Limits
shbg_assay_values <- c(1, 2, 3, 4, 5, 6, 7, 9)
shbg_assay_tech_max <- c(300, 300, 180, 180, 1000, 180, 180, 180)
shbg_assay_tech_min <- c(5, 5, 1, 0.2, 0, 1, 1, 1)
shbg_assay_log_max <- c(100, 100, 100, 100,100,100,100,100)
shbg_assay_log_min <- c(50, 50, 50, 50, 50, 50, 50, 50)
# DHEAS Limits
dheas_assay_values <- c(1, 2, 3, 4, 5)
dheas_assay_tech_max <- c(10000, 10000, 10000, 10000, 10000)
dheas_assay_tech_min <- c(50, 150, 150, 150, 150)
dheas_assay_log_max <- c(6000, 6000, 6000, 6000, 6000)
dheas_assay_log_min <- c(250, 250, 250, 250, 250)
# LH Limits
lh_assay_values <- c(1, 2, 3, 4, 5, 6)
lh_assay_tech_max <- c(300, 300, 100, 200, 100, 100)
lh_assay_tech_min <- c(1, 1, 0.2, 0.1, 0.2, 0.2)
lh_assay_log_max <- c(60, 60, 60, 60, 60, 60)
lh_assay_log_min <- c(2, 2, 2, 2, 2, 2)
# FSH Limits
fsh_assay_values <- c(1, 2, 3, 4, 5, 6)
fsh_assay_tech_max <- c(450, 450, 100, 170, 170, 170)
fsh_assay_tech_min <- c(1.5, 1.5, 1, 0.1, 0.1, 0.1)
fsh_assay_log_max <- c(40, 40, 40, 40, 40, 40)
fsh_assay_log_min <- c(2, 2, 2, 2, 2, 2)
# Fasting Glucose Limits
g0_assay_values <- c(1,10,2,3,4,5,6,7,8)
g0_assay_tech_max <- c(600, 450, 450, 450, 450, 540, 500, 450, 900)
g0_assay_tech_min <- c(19, 10, 10, 10, 10, 1, 70, 10, 10)
g0_assay_log_max <- c(200, 200, 200, 200, 200, 200, 200, 200, 200)
g0_assay_log_min <- c(50, 50, 50, 50, 50, 50, 50, 50, 50)
# Fasting Insulin Limits
i0_assay_values <- c(1,2,3,4,5,6,7)
i0_assay_tech_max <- c(350, 441, 300, 300, 300, 441, 300)
i0_assay_tech_min <- c(5, 2.9, 2, 2, 2, 2.9, 2)
i0_assay_log_max <- c(120,120, 120, 120, 120, 120, 120)
i0_assay_log_min <- c(6, 6, 6, 6, 6, 6, 6)

################################################
# Read in data and filtering                   #
################################################

## Read in input file (must have columns according to cluster_input_cols)
m.df0 <- read.delim(input_file, na.strings=c("", "#N/A", "missing"))
str(m.df0)
```

```{r, echo=FALSE}

## Remove duplicate entries (keep the first occurrence for each set of duplicates)
dupidx = duplicated(m.df0[, cluster_input_cols[1:(3+length(traitnames))]])
if(sum(dupidx)>0) warning(paste("Removing", sum(dupidx), "duplicates from data"))
m.df <- m.df0[!dupidx,]




## Only include data with non-zero data for age and the 8 traits
#zeroidx = !is.na(rowSums(m.df[, c(age, bmi, traitnames)]==0)) > 0
#if(sum(zeroidx)>0) {
#    warning(paste("Removing", sum(zeroidx), "records due to zero values"))
#    for(ii in c(age, bmi, traitnames)) {
#        tmpidx = m.df[, ii]==0
#        if(sum(tmpidx)>0) message(paste(ii, "has", sum(tmpidx), "zero values"))
#    }
#    m.df <- m.df[!zeroidx, ]

## Calculate z-scores, adjusting by trait to determine the cutoffs for SD outliers by trait
for (var in traitnames) {
  method <- paste0(var, assay)
  z <- paste(var, 'z.flag', sep=d)
  m.df[,method] <- as.factor(as.character(m.df[,method]))
  m.df[, z] <- ave(m.df[,var], m.df[,method], FUN = scale)
}

# Creating the variables for the flags for z scores for each trait
traitflags = c()
for (var in traitnames) {
  names <- c(paste0(var, '.z.flag', sep=''))
  traitflags[[names]] <- names
}

# Z-scores with SD > 4 will be flagged == 1, if not, value in flag will be == 0
traitflags <- as.character(traitflags)
for (var in traitflags) {
  m.df[,var] <- ifelse(m.df[,var] > 4, 1, 0)
}

# Create variables for technical limit flags and set them == 0
trait.tech.flags = c()
for (var in traitnames) {
  names <- c(paste0(var, '_tech_flag', sep=''))
  trait.tech.flags[[names]] <- names
}
trait.tech.flags <- as.character(trait.tech.flags)
for (var in trait.tech.flags) {
  m.df[,var] <- 0
}

# Create variables for logical limit flags and set them == 0
trait.log.flags = c()
for (var in traitnames) {
  names <- c(paste0(var, '_log_flag', sep=''))
  trait.log.flags[[names]] <- names
}
trait.log.flags <- as.character(trait.log.flags)
for (var in trait.log.flags) {
  m.df[,var] <- 0
}

# Create new dataframe for each assay method to hold the technical and logical limits
for (var in traitnames) {
  newlist <- lapply(list(get(paste0(var,"_assay_tech_max")), get(paste0(var,"_assay_tech_min")), get(paste0(var,"_assay_log_max")), get(paste0(var,"_assay_log_min"))), setNames, get(paste0(var,"_assay_values")))
  newdf <- as.data.frame(do.call(rbind, newlist))
  rownames(newdf) <- c("Technical_max", "Technical_Min", "Logical_Max", "Logical_Min")
  assign(paste(var,'.ranges',sep=''),newdf)
}
```

## Technical and Logical Range Checks

### Testosterone
```{r eda_range_t, echo=FALSE, error=FALSE, message=TRUE, out.width="50%"}

# Testosterone - Technical and Logical Range Checks
for (i in 1:nrow(m.df)) {
  if(is.na(m.df[[i,"T_assay_method"]])) {
    m.df[[i,"T_tech_flag"]] <- 0
    m.df[[i,"T_log_flag"]] <- 0
  }
  else for (j in colnames(T.ranges)) {
    if (m.df[[i,"T_assay_method"]] == as.numeric(j)) {
      if (m.df[[i,"T"]] > T.ranges[[1,j]]) {
        m.df[[i,"T_tech_flag"]] <- 1
      }
      if (m.df[[i,"T"]] < T.ranges[[2,j]]) {
        m.df[[i,"T_tech_flag"]] <- 1
      }
      if (m.df[[i,"T"]] > T.ranges[[3,j]]) {
        m.df[[i,"T_log_flag"]] <- 1
      }
      if (m.df[[i,"T"]] < T.ranges[[4,j]]) {
        m.df[[i,"T_log_flag"]] <- 1
    }
    }
  }
}

datatable(T.ranges, options = list(dom = 't', columnDefs = list(list(className = 'dt-center', targets = "_all"))))

tmp <- m.df[m.df$T_tech_flag == 1,]
tmp <- tmp[,c("sample_id", "T", "T_assay_method")]

if(nrow(tmp) > 0) {
 message(paste("\n\nThere are", length(which(m.df$T_tech_flag == 1)), "values that are outside of Testosterone technical limits"))
  datatable(tmp, rownames = FALSE, options = list(searching = FALSE, columnDefs = list(list(className = 'dt-center', targets = "_all"))))
} else 
   message("\n\nThere are no values outside of the testosterone technical limits")


tmp <- m.df[m.df$T_log_flag == 1,]
tmp <- tmp[,c("sample_id", "T", "T_assay_method")]
if(nrow(tmp) > 0) {
 message(paste("\n\nThere are", length(which(m.df$T_log_flag == 1)), "values that are outside of Testosterone logical limits"))
  datatable(tmp, rownames = FALSE, options = list(searching = FALSE, columnDefs = list(list(className = 'dt-center', targets = "_all"))))
} else 
   message("\n\nThere are no values outside of the testosterone logical limits")


```

